DRAFT: Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-03-12 products/manuscript/Manuscript.qmd
manuscript update
saving manuscript edits. began restructuring the sections.

This project will lay the foundation for my dissertation research. **The
goal for this phase of the project is to answer which environmental,
ecological, and temporal factors influence the transmission cycles for
EEE in southeastern Massachusetts.**

@@

## General Background Information

@@

Since 2000, there have been 45 human cases recorded in Massachusetts,
resulting in 22 deaths. There are no human vaccines for EEE, and
treatment consists of palliative care only. Prevention measures include
personal behaviors to avoid mosquito bites and decreasing the mosquito
population through pesticide use and environmental modifications like
removing standing water. The Bristol County Mosquito Control Project
(BCMCP) coordinates mosquito surveillance and testing in the county from
June to October. BCMCP has used the same sentinel collection sites and
trapping methods for over 20 years. Once trapped, mosquitoes are sorted
by species, and vector species are submitted to the Massachusetts
Department of Health’s (MDPH) State Lab for PCR virus testing. Results
are available within 24 hours of submission. When rates of EEE mosquito
infections are above a defined threshold, MDPH notifies the local boards
of health and recommends preventive measures. Recommended measures
include outreach and education to increase personal protective practices
. Additionally, MDPH may recommend that cities and towns in the affected
areas cancel evening outdoor events and discourage outdoor activities
when mosquito vectors are most active (dawn and dusk). When infection
rates are high enough to indicate imminent human transmission, the
Commonwealth of Massachusetts will recommend and fund aerial pesticide
applications over the affected areas.

## Description of data and data source

*The cornerstone of this research will be the mosquito collection data
provided by a mosquito control project in SE Massachusetts. A formal
data-sharing agreement will be in place by January 2024. The dataset
encompasses nearly two decades of detailed mosquito surveillance. The
datasets include spatial identifiers. Mosquito collection data will be
integrated with other datasets using spatial or temporal attributes.
These datasets will include biological and ecological data,
environmental and spatial data, and demographic data, which are crucial
in understanding the transmission dynamics of EEE. Potential datasets
with their sources are outlined in the table below. Other datasets may
be included based on availability and need.*

| Data                         | Description                                                                                                                                           | Source                                |
|--------------|--------------------------------------------|--------------|
| Mosquito Surveillance Data   | Records of mosquito species counts, locations, and dates of collection. Data on the presence of EEE in mosquito populations is particularly valuable. | XXX County Mosquito Control           |
| Weather Data                 | Temperature, humidity, rainfall, and wind speed influence mosquito activity, population dynamics, and virus transmission.                             | NOAA                                  |
| Bird Population Data         | Since birds are a natural reservoir for EEE, information on bird populations and migration patterns could be relevant.                                | eBird                                 |
| Human Case Reports           | Data on confirmed human cases of EEE, including location, date of onset, and clinical outcomes.                                                       | MA Dept of Public Health publications |
| Veterinary Surveillance Data | Since horses and other mammals can also be affected by EEE, veterinary records could provide early warning signs of virus activity in an area.        | MA Dept of Public Health              |

@@

\*This project will lay the foundation for my dissertation research.
**The goal for this phase of the project is to answer which
environmental, ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts.** While the
following questions will not be addressed in this phase, this project
will hopefully inform research that will answer the following questions:

>>>>

# Problem statement

Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts and which
models best describe the relationship between these factors.

This project will lay the foundation for dissertation research that will
begin in June, 2024. The research will focus on three

@@

## Background

@@

Since 2000, there have been 45 human cases recorded in Massachusetts,
resulting in 22 deaths. There are no human vaccines for EEE, and
treatment consists of palliative care only. Prevention measures include
personal behaviors to avoid mosquito bites and decreasing the mosquito
population through pesticide use and environmental modifications like
removing standing water.

The Bristol County Mosquito Control Project (BCMCP) coordinates mosquito
surveillance and testing in the county from June to October. BCMCP has
used the same sentinel collection sites and trapping methods for over 20
years. Once trapped, mosquitoes are sorted by species, and vector
species are submitted to the Massachusetts Department of Health’s (MDPH)
State Lab for PCR virus testing. Results are available within 24 hours
of submission. When rates of EEE mosquito infections are above a defined
threshold, MDPH notifies the local boards of health and recommends
preventive measures. Recommended measures include outreach and education
to increase personal protective practices . Additionally, MDPH may
recommend that cities and towns in the affected areas cancel evening
outdoor events and discourage outdoor activities when mosquito vectors
are most active (dawn and dusk). When infection rates are high enough to
indicate imminent human transmission, the Commonwealth of Massachusetts
will recommend and fund aerial pesticide applications over the affected
areas.

## Study area

Bristol County, Massachusetts covers an area of 691 square miles with a
population density of 962 people per square mile. It borders Rhode
Island to the west, Plymouth County, MA to the east, and Norfolk County,
MA to the north.The county is situated on Buzzard's Bay to the south.

Atlantic White Cedar (AWC) swamps span over 4,300 acres within
Massachusetts, primarily concentrated in Plymouth and Bristol counties.
These swamps serve as critical breeding habitats for Culiseta melanura
(MEL), which is recognized as the primary enzootic vector for Eastern
Equine Encephalitis (EEE). AWC trees have large roots that form "crypts"
that hold water throughout the year. These crypts serve as a main
habitat for MEL larval growth.

Between 2003 and 2022 Plymouth and Bristol counties were the epicenters
for EEE transmission in Massachusetts, accounting for 24 out of the
state's 41 human EEE cases.The Bristol County Mosquito Control Project
(BCMCP) conducts mosquito surveillance and abatement activities for the
county and often coordinates activities with Plymouth County Mosquito
Control Project.

## Data collection and processing

The cornerstone of this project is mosquito surveillance data. This
dataset includes mosquito counts by species for all mosquito traps set
in Bristol County, MA between 2007 to 2023. In addition to mosquito
counts by species, the dataset includes the town or city in which they
were collected, the collection data, and whether the mosquitoes were
submitted for virus testing.

While the surveillance data provide insight into mosquito population
estimates and changes in species composition, it is missing one key
feature. The virus isolation results are not included in this dataset.
However, virus isolations in mosquitoes are published on the MDPH
website by town/city and collection date. With this information, one can
identify the trap event contained in the dataset which allow for
calculations of key measure, such as minimum infection rate (MIR).

The surveillance dataset is structured to conform with state and federal
reporting requirements. As a result, the format of the data must be
transformed to in order to make it ready for analysis. Once the data
structure is corrected, it can be joined to other relevant datasets
using spatial or temporal attributes.

This dataset will be integrated with other datasets using spatial or
temporal attributes. These datasets will include biological and
ecological data, environmental and spatial data, and demographic data,
which are crucial in understanding the transmission dynamics of EEE.
Potential datasets with their sources are outlined in the table below.
Other datasets may be included based on availability and need.

| Data                         | Description                                                                                                                                           | Source                                |
|-------------------|----------------------------------|-------------------|
| Mosquito Surveillance Data   | Records of mosquito species counts, locations, and dates of collection. Data on the presence of EEE in mosquito populations is particularly valuable. | XXX County Mosquito Control           |
| Weather Data                 | Temperature, humidity, rainfall, and wind speed influence mosquito activity, population dynamics, and virus transmission.                             | NOAA                                  |
| Bird Population Data         | Since birds are a natural reservoir for EEE, information on bird populations and migration patterns could be relevant.                                | eBird                                 |
| Human Case Reports           | Data on confirmed human cases of EEE, including location, date of onset, and clinical outcomes.                                                       | MA Dept of Public Health publications |
| Veterinary Surveillance Data | Since horses and other mammals can also be affected by EEE, veterinary records could provide early warning signs of virus activity in an area.        | MA Dept of Public Health              |

@@

\*This project will lay the foundation for my dissertation research.
**The goal for this phase of the project is to answer which
environmental, ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts.**

While the following questions will not be addressed in this phase, this
project will hopefully inform research that will answer the following
questions:
